paper2_resistance_mechanisms_mixtures

### 2017-02-22 inst/documents/paper2/paper2_resistance_mechanisms_mixtures.Rmd
paper2 writing

```{r, eval=FALSE, include=FALSE}
#!!!! paper protocol
#!!!! might be easier to put text & figs into separate Rmd files after draft stage
#!!!! drafts knit to pdf with pdf_document in YAML and this chunk eval=FALSE
#!!!! submission MS, knit to word_document, YAML, this chunk eval=FALSE, 
#!!!!  in WORD: delete figs, add fig numbers to fig legends, set double space, add line numbers 
#!!!! submission figs, set this chunk to eval=TRUE

@@

THIS IS A CONFIDENTIAL COPY OF THE MS. Please do not circulate without the express permission of one of the authors. Comments and & suggestions welcome

@@

_must be 350 words, currently 375 with subheadings Background, Results & Conclusions_

@@

_TODO add in here about theory of IR & protocols for resistance management._

@@

The simulation represents a population of randomly mixing individuals using standard population genetic approaches to avoid the need to follow every individual. One locus is represented for each insecticide. There are two potential alleles per locus. Each allele confers either resistance or susceptibility to the insecticide. Fitness is the main currency of the model, representing how much each genotype survives and reproduces, with a maximum value of 1 defined as the fitness of the SS genotype in absence of the insecticide. The fitness of an individual depends on it's alleles, exposure to the insecticide and other inputs that can be set in the model. These inputs include the effectiveness of the insecticide, the dominance of the resistance allele and the ability of resistance to restore fitness in the presence of the insecticide (see Table 1). Each of these inputs can be varied as a proportion between 0 and 1. We use 'Resistance restoration' to represent how much the RR genotype restores fitness in the presence of the insecticide. The more conventional 'selection coefficient' is calculated by multiplying Resistance restoration by insecticide effectiveness. Doing it this way allows us to vary all inputs between 0 and 1 without fitness going above 1 in the model.

The simulation proceeds through generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness. This acts to make the fitter alleles more common over time. Thus a summary of the selection stage is :

> selection coefficient = resistance restoration * effectiveness  
>
> Locus fitness : 	
> SS = 1-effectiveness  
> RS = 1-effectiveness + (dominance*selection coeff.)  
> RR = 1-effectiveness + selection coeff.  
> 
> Genotype fitness :   
> SS1SS2 = SS * SS  
> SS1RS2 = SS * RS  
> … and for other combinations  
> RR1RR2 = RR * RR  
> 
> Population fitness = Genotype fitness * exposure to insecticide  
> 
> In each generation :   
> Allele frequency[t+1] = Allele frequency[t] * Population fitness

@@

The development of insecticide resistance in operational field settings is subject to considerable uncertainty [@Ransom2016]. The model we present here is able to cover the range of that uncertainty, yet in this paper, to develop an understanding of the key mechanisms we restrict our explorations to a limited region of paramter space (e.g. holding most inputs constant at intermediate values while modifying single inputs in isolation). Thus there is the potential that the system may behave differently in the field than we describe here. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. In addition we provide an online user interface to the model [@South2017a] and enabling readers to investigate model behaviour in areas of parameter space we have not shown here.

@@

![New figure showing how fitness determined by inputs](paper2_mod_graphic_a_to_c.png)

![Screenshot of one online model user interface, accessible at : https://andysouth.shinyapps.io/resistmob2/. The user can modify values of the input parameters considered in this paper using simple sliders and run the model to get graphs of resulting resistance frequency over time. Two scenarios (A&B) can be run and the results viewed side by side. This makes it easy to explore the effect of changing individual inputs.](paper2_UI1.png)

@@

  print( ggplot(df_resist, aes(x=generation, y=resistance, colour=factor(P_1))) + 
         theme_bw() +
         theme(legend.position = "bottom", legend.key = element_blank()) + 
         guides(colour = guide_legend(reverse=TRUE)) + 
         labs(colour = "start\nfrequency") +
         coord_trans(y = "log10") + 
         geom_line()

       )
```

```{r Fig5, eval=TRUE, echo=FALSE, fig.height=7, fig.width=7, fig.show='hold', fig.align='left', fig.cap="Influence of insecticide effectiveness and exposure on time-to-resistance for mixtures and sequences. Exposure to the insecticides is increased from row 1 (A&B) to row 2 (C&D). The effectiveness of one insecticide is increased from column 1 (A&C) to column2 (B&D). On the upper X axis s&m indicate where the 50% resistance threshold is reached for the sequence and the mixture. In the lower right of each panel the the ratio of time-to-resistance for mixture/sequence is shown rounded to 1 decimal place to give an indication of the relative performance of mixtures and sequences.  A. All control inputs equal at 0.5 : time-to-resistance is longer for sequential use, B. Effectiveness of insecticide1 increased from 0.5 to 0.8 : time-to-resistance is longer for the mixture, C. Exposure increased to 0.8 : time-to-resistance is longer for sequential use, D. Effectiveness of insecticide1 and exposure increased to 0.8 : time-to-resistance equal for mixture and sequence. Increasing effectiveness increases times-to-resistance for mixtures and improves their performance relative to sequences. Increasing exposure decreases times-to-resistance for mixtures and reduces their performance relative to sequences. "}

@@

#eff1 & expos 0.8
runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.5 , h.RS2_0B = 0.5 , exposure = 0.8 , phi.SS1_A0 = 0.8 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('s','','a','m'), ylab="", ylabs = FALSE, cex.axis = 0.8, addLegend=FALSE, main='', maxX = 120, labelMixSeqRatio = 1 )

@@

```{r Fig6, eval=TRUE, echo=FALSE, fig.height=7, fig.width=7, fig.show='hold', fig.align='left', fig.cap="Influence of the effectiveness of both insecticides on time-to-resistance for mixtures and sequences. Effectiveness of insecticide 2 is increased from row 1 (A&B) to row 2 (C&D). The effectiveness of insecticide 1 is increased from column 1 (A&C) to column2 (B&D). s&m on the upper X axis and mix/seq in th lower right of each panel are explained in the legend to Fig 5. A. Effectiveness of insecticide1 increased from 0.5 to 0.8, B. Effectiveness of insecticide1 increased from 0.5 to 1, C. Effectiveness of insecticides 1 and 2 increased to 0.8, D. Effectiveness of insecticide2 0.8 and of insecticide1 1. With the effectiveness of at least one insecticide greater than or equal to 0.8, times-to-resistance are longer for the mixture in all scenarios."}

@@

```{r Fig7, eval=TRUE, echo=FALSE, fig.height=7, fig.width=7, fig.show='hold', fig.align='left', fig.cap="Influence of dominance and resistance restoration on time-to-resistance for mixtures and sequences. Dominance of the allele coding for resistance to insecticide 1 is increased from row 1 (A&B) to row 2 (C&D). Resistance restoration for the allele coding for resistance to insecticide 1 is increased from column 1 (A&C) to column2 (B&D). s&m on the upper X axis and mix/seq in th lower right of each panel are explained in the legend to Fig 5. A. All control inputs equal at 0.5, B. Resistance restoration for insecticide1 increased from 0.5 to 0.8, C. Dominance for insecticide1 increased to 0.8, D. Resistance restoration and dominance for insecticide1 increased to 0.8. Changing dominance and resistance restoration does not change the relative ordering of mixtures and sequences, time-to-resistance remains longest for sequences in all 4 scenarios."}

@@

```{r Fig8, eval=TRUE, echo=FALSE, fig.height=7, fig.width=7, fig.show='hold', fig.align='left', fig.cap="Influence of starting frequencies of resistance on time-to-resistance for mixtures and sequences. Starting frequency of the gene conferring resistance to insecticide 1 is decreased from row 1 (A&B) to row 2 (C&D). Effectiveness of insecticide 1 is increased from column 1 (A&C) to column2 (B&D). s&m on the upper X axis and mix/seq in th lower right of each panel are explained in the legend to Fig 5. A. All control inputs equal at 0.5, starting frequencies of resistance at 0.01, B. Effectiveness for insecticide1 increased from 0.5 to 0.8, C. Starting frequency of resistance for insecticide1 decreased from 0.01 to 0.001, D. Effectiveness for insecticide1 increased from 0.5 to 0.8 and starting frequency for insecticide1 decreased from 0.01 to 0.001. In these scenarios the starting frequencies do not change better performance of sequences at low effectiveness and mixtures at high effectiveness."}

>>>>

```{r, eval=FALSE, include=FALSE}
#!!!! paper protocol
#!!!! might be easier to put text and figs into separate Rmd files after draft stage
#!!!! drafts knit to pdf with pdf_document in YAML and this chunk eval=FALSE
#!!!! submission MS, knit to word_document, YAML, this chunk eval=FALSE, 
#!!!!  in WORD: delete figs, add fig numbers to fig legends, set double space, add line numbers 
#!!!! submission figs, set this chunk to eval=TRUE

@@

THIS IS A CONFIDENTIAL COPY OF THE MS. Please do not circulate without the express permission of one of the authors. Comments and suggestions welcome

@@

_must be 350 words, currently 375 with subheadings Background, Results and Conclusions_

@@

_TODO add in here about theory of IR and protocols for resistance management._

@@

The simulation represents a population of randomly mixing individuals using standard population genetic approaches to avoid the need to follow every individual. One locus is represented for each insecticide. There are two potential alleles per locus. Each allele confers either resistance or susceptibility to the insecticide. Fitness is the main currency of the model, representing how much each genotype survives and reproduces, with a maximum value of 1 defined as the fitness of the SS genotype in absence of the insecticide. The fitness of an individual depends on it's alleles, exposure to the insecticide and other inputs that can be set in the model. These inputs include the effectiveness of the insecticide, the dominance of the resistance allele and the ability of resistance to restore fitness in the presence of the insecticide (see Table 1).

The calculation of fitness for each genotype when considering just a single insecticide is illustrated in Figure 1. In the model itself this is repeated for both insecticides and the fitnesses multiplied. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS and resistance restoration 'restores' a portion of the fitness for RR. A value of 1 for resistance restoration would bring the fitness of RR back to 1 (equivalent to the unexposed SS), a value of 0 for resistance restoration would lead to the RR having the same fitness as SS in the presence of the insecticide and effectively there would be no resistance. Dominance of restoration determines how the fitness for SR sits between that of SS and RR for those that are exposed to the insecticide. For those that are not exposed, fitness of SS is set to 1 by definition and resistance cost determines the fitness of RR. Dominance of cost determines how the fitness for SR sits between that of SS and RR in the absence of the insecticide (Figure 1).

Each of these inputs can be varied as a proportion between 0 and 1. We use 'Resistance restoration' to represent how much the RR genotype restores fitness in the presence of the insecticide. The more conventional 'selection coefficient' is calculated by multiplying resistance restoration by insecticide effectiveness.

The simulation proceeds through generations. In each generation selection is represented by multiplying genotype frequencies in the population by their relative fitness. This acts to make the fitter alleles more common over time. Thus a summary of the selection stage is as follows with 1 & 2 representing the two insecticides :

> selection coefficient = resistance restoration * effectiveness  
>
> Single gene, insecticide fitness : 	
> SS = 1-effectiveness  
> RS = 1-effectiveness + (dominance*selection coeff.)  
> RR = 1-effectiveness + selection coeff.  
> 
> Double gene, insecticide fitness :   
> SS1SS2 = SS1 * SS2  
> SS1RS2 = SS1 * RS2  
> … and for other combinations  
> RR1RR2 = RR1 * RR2  
> 
> Population fitness = Double fitness * exposure to insecticides  
> 
> In each generation :   
> Allele frequency[t+1] = Allele frequency[t] * Population fitness

@@

The development of insecticide resistance in operational field settings is subject to considerable uncertainty [@Ranson2016]. The model we present here is able to cover the range of that uncertainty, yet in this paper, to develop an understanding of the key mechanisms we restrict our explorations to a limited region of paramter space (e.g. holding most inputs constant at intermediate values while modifying single inputs in isolation). Thus there is the potential that the system may behave differently in the field than we describe here. However, the results presented here are supported by an earlier analysis where we ran 10,000 unique scenarios varying more inputs [@Levick2017]. In addition we provide an online user interface to the model [@South2017a] and enabling readers to investigate model behaviour in areas of parameter space we have not shown here.

@@

```{r Fig1, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=7.5, fig.show='hold', fig.align='left', fig.cap="Diagrammatic representation of the effect of model inputs on the fitness of genotypes for a single insecticide. Fitness is shown on the y-axis and the different genotypes (SS, SR, RR) on the x axis. Firstly the exposure input determines the proportion of the population in the left and right panels (exposed and not exposed). For those that are exposed (left panel) insecticide effectiveness sets the fitness for SS, resistance restoration 'restores' a portion of the fitness for RR and dominance of restoration determines how the fitness for SR sits between that of SS and RR. For those that are not exposed, fitness of SS is set to 1 by definition, resistance cost determines the fitness of RR and again dominance of cost determines how the fitness for SR sits between that of SS and RR. In this example effectiveness=0.8, dominance of restoration=0.4, resistance restoration=0.5, dominance of cost=0.4 and resistance cost=0.3."}

plot_fit_calc( effectiveness=0.8, resistance_restoration=0.5, dominance_restoration=0.4, dominance_cost=0.4, cost=0.3 )

![Screenshot of one online model user interface, accessible at : <https://andysouth.shinyapps.io/resistmob2/>. The user can modify values of the input parameters considered in this paper using simple sliders and run the model to get graphs of resulting resistance frequency over time. Two scenarios (A and B) can be run and the results viewed side by side. This makes it easy to explore the effect of changing individual inputs.](paper2_UI1.png)

@@

  print( ggplot(df_resist, aes(x=generation, y=resistance, colour=factor(P_1))) + 
         theme_bw() +
         theme(legend.position = "bottom", legend.key = element_blank()) + 
         guides(colour = guide_legend(reverse=TRUE)) + 
         labs(colour = "start\nfrequency") +
         coord_trans(y = "log10") + 
         geom_line()   
       )

```{r Fig5, eval=TRUE, echo=FALSE, fig.height=7, fig.width=7, fig.show='hold', fig.align='left', fig.cap="Influence of insecticide effectiveness and exposure on time-to-resistance for mixtures and sequences. Exposure to the insecticides is increased from row 1 (A and B) to row 2 (C and D). The effectiveness of one insecticide is increased from column 1 (A and C) to column2 (B and D). On the upper X axis 's' and 'm' indicate where the 50 percent resistance threshold is reached for the sequence and the mixture. In the lower right of each panel the the ratio of time-to-resistance for mixture/sequence is shown rounded to 1 decimal place to give an indication of the relative performance of mixtures and sequences.  A. All control inputs equal at 0.5 : time-to-resistance is longer for sequential use, B. Effectiveness of insecticide1 increased from 0.5 to 0.8 : time-to-resistance is longer for the mixture, C. Exposure increased to 0.8 : time-to-resistance is longer for sequential use, D. Effectiveness of insecticide1 and exposure increased to 0.8 : time-to-resistance equal for mixture and sequence. Increasing effectiveness increases times-to-resistance for mixtures and improves their performance relative to sequences. Increasing exposure decreases times-to-resistance for mixtures and reduces their performance relative to sequences. "}

@@

#eff1  and  expos 0.8
runcurtis_f2( max_gen=500,  P_1 = 0.01 , P_2 = 0.01 , h.RS1_A0 = 0.5 , h.RS2_0B = 0.5 , exposure = 0.8 , phi.SS1_A0 = 0.8 , phi.SS2_0B = 0.5 , rr_restoration_ins1 = 0.5 , rr_restoration_ins2 = 0.5 , addCombinedStrategy = FALSE, strategyLabels = c('s','','a','m'), ylab="", ylabs = FALSE, cex.axis = 0.8, addLegend=FALSE, main='', maxX = 120, labelMixSeqRatio = 1 )

@@

```{r Fig6, eval=TRUE, echo=FALSE, fig.height=7, fig.width=7, fig.show='hold', fig.align='left', fig.cap="Influence of the effectiveness of both insecticides on time-to-resistance for mixtures and sequences. Effectiveness of insecticide 2 is increased from row 1 (A and B) to row 2 (C and D). The effectiveness of insecticide 1 is increased from column 1 (A and C) to column2 (B and D). The 's' and 'm' on the upper X axis and 'mix/seq' in the lower right of each panel stand for sequence and mixture and are explained in the legend to Fig 5. A. Effectiveness of insecticide1 increased from 0.5 to 0.8, B. Effectiveness of insecticide1 increased from 0.5 to 1, C. Effectiveness of insecticides 1 and 2 increased to 0.8, D. Effectiveness of insecticide2 0.8 and of insecticide1 1. With the effectiveness of at least one insecticide greater than or equal to 0.8, times-to-resistance are longer for the mixture in all scenarios."}

@@

```{r Fig7, eval=TRUE, echo=FALSE, fig.height=7, fig.width=7, fig.show='hold', fig.align='left', fig.cap="Influence of dominance and resistance restoration on time-to-resistance for mixtures and sequences. Dominance of the allele coding for resistance to insecticide 1 is increased from row 1 (A and B) to row 2 (C and D). Resistance restoration for the allele coding for resistance to insecticide 1 is increased from column 1 (A and C) to column2 (B and D). The 's' and 'm' on the upper X axis and mix/seq in the lower right of each panel stand for sequence and mixture and are explained in the legend to Fig 5. A. All control inputs equal at 0.5, B. Resistance restoration for insecticide1 increased from 0.5 to 0.8, C. Dominance for insecticide1 increased to 0.8, D. Resistance restoration and dominance for insecticide1 increased to 0.8. Changing dominance and resistance restoration does not change the relative ordering of mixtures and sequences, time-to-resistance remains longest for sequences in all 4 scenarios."}

@@

```{r Fig8, eval=TRUE, echo=FALSE, fig.height=7, fig.width=7, fig.show='hold', fig.align='left', fig.cap="Influence of starting frequencies of resistance on time-to-resistance for mixtures and sequences. Starting frequency of the gene conferring resistance to insecticide 1 is decreased from row 1 (A and B) to row 2 (C and D). Effectiveness of insecticide 1 is increased from column 1 (A and C) to column2 (B and D). The 's' and 'm' on the upper X axis and mix/seq in the lower right of each panel stand for sequence and mixture and are explained in the legend to Fig 5. A. All control inputs equal at 0.5, starting frequencies of resistance at 0.01, B. Effectiveness for insecticide1 increased from 0.5 to 0.8, C. Starting frequency of resistance for insecticide1 decreased from 0.01 to 0.001, D. Effectiveness for insecticide1 increased from 0.5 to 0.8 and starting frequency for insecticide1 decreased from 0.01 to 0.001. In these scenarios the starting frequencies do not change better performance of sequences at low effectiveness and mixtures at high effectiveness."}
